---
title: "R programming KT2"
output: html_document
---
*Vectors*
Vectors are 1 dimensional arrays of a single type of data. each element in a vector can have a name, these vectors can also be used to vectorize operations to save time make the program efficient. When different datatypes are used R automatically tries to convert all the elements to a single type, if it fails then it returns an error
  *Creating Vectors*
    a)Using the ':' operator to generate a vector
    ```{r}
    Vector1 = -8:10
    Vector1
    ```
    b) Using the seq() function
    ```{r}
    Vector1 = seq(10,20,0.5)
    Vector1
    ```
    c) Using the rep() function
    ```{r}
    Vector1 = rep(2:4,5)
    Vector1
    ```
    d) Using the c() function
    ```{r}
    Vector1 = c("a","b","c","d","e", 7, F, 8L)
    Vector1
    ```
  
  *Accessing elements in a vector and slicing a vector*
    We can access elements in a vector by passing the inde of an element in the vector.It can be an integer or a vector, R also accepts negative arguments. We can also use the names of individual elements to access a vector. Passing an array of booleans can also filter elements as needed.
    ```{r}
    Vector1[3]
    
    names(Vector1) = c("a","b","c","d","e","f","g","h")
    
    Vector1["c"]
    
    Vector2 = seq(10,15,length.out= 7)
    Vector2
    
    Vector2[Vector2<13.5]
    ```
  
  *Performing operations with a vector*
  Instead of using loops and complicated functions to perform basic operations with a vector, we can just use mathematical operators to perform the said operators.
```{r}
    a = c(3, 6, 4, 7, 3, 5)
    b = 1L:6L
    c = rep(c(T,F), 3)
    d = c(2,2,2)
    
    a-b
    
    b*c
    
    #Recycling the vector
    b+d
```
  
  *Adding and removing elements from a vector*
  To add or remove an element we just need to pass the inde of the new element
```{r}
  b[7] = 7
  b[7] = NULL
```

*Lists*
Lists are general purpose vectors that can store elements of multiple data types or even store other lists , vectors and matrices
  *Creating Lists*
```{r}
    List1 = list("Hello World",3 + 5i, 1L:4L, FALSE, "Goodbye World", matrix(1:8, nrow = 2, ncol = 4))
    List1 
```
  
  *Accessing elements in a list*
```{r}
    List1[3]
    #Giving names to all the elements in a list
    names(List1) = c("Hello", "ComplexNumber","Logical","IntVector","Goodbye","Mat")
    
    List1["IntVector"]
    List1$Hello
```
 *Adding and removing elements from a list*
  To add or remove an element we just need to pass the inde of the new element
```{r}
    List1[length(List1)+1] = "NOoooOOooOOoooOOoooooo"
    List1[length(List1)+1] = NULL
```

*Matrices*
Matrices are 2 dimensional vectors.
  *Creating a Matrix*
    a) Using the matrix() method
    ```{r}
    mat1 = matrix(1:9, nrow = 3)
    mat1
    mat1 = matrix(1:9, ncol = 3,byrow = TRUE)
    mat1
    ```
    b) Using rbind()
    ```{r}
    mat1 = rbind(a,b,c,seq(-5,-15, length.out = 6))
    mat1
    ```
    c) Using cbind()
    ```{r}
    mat1 = cbind(a,b,c,seq(5,15, length.out = 6))
    mat1
    ```
    
  *Accessing/Slicing/Subsetting from a matrix*
  It is very similar to accessing from a vector.
    ```{r}
    mat1 = t(mat1)
    mat1
    
    mat1[3,]
    
    mat1[,4]
    
    mat1[3,4]
    #Giving row names and column names to Mat1
    rownames(mat1) = c("Monday","Tuesday","Wedneday","Thursday")
    colnames(mat1) = c("Maths", "Science", "CS", "English", "Hindi", "Tamil")
    mat1
    
    mat1["Monday", "Science",]
  
    mat1[mat1[,"Science"] %% 2 == 0,"Science"]
         

    
    
    mat1[1:3,4:6]
    ```   
    
  *Operations with matrices*
  The operations on matrices are also very similar to operations on vectors.
```{r}
     mat2 = rbind(a+b,c*b,b/a, rep(c(1,2),3))
     mat2 = round(mat2)
     
     mat1 * mat1
     mat1 + mat2
     mat1 - mat2
     mat1 / mat2
     
     #matrix multiplication
      mat1 %*% t(mat2)
      ?rbind()
      
```
  *Adding/Removing elements and Merging matrices*
  Add,Remove or Merge a matrix we can use either rbind() or the cbind function, the choice of function determines weather the matrices are merged along its rows or columns
```{r}
    rbind(mat1,mat2)
    cbind(mat1,mat2)
```




