---
title: "R Programming KT1"
output: html_document
---
                           *Introduction to R programming*
 
R is a programming language primarily used by Data Scientists/Analysts to help perform  various operations on large data sets using the powerful modules built for R.

*Variables:*
R variable names can only contain alphanumeric characters, underscore and period. These variable names can only start with a period or an alphabet character. The variable names are also case sensitive.

   *Examples of valid variable names:*
      Variable1
      .Variable1
      .Variable_1 

   *Examples of invalid variable names:*
      _Variable1
      V@riable!
      1Variable
      

*Assigning a Variable:*
```{r}
      Variable1 <- "Hello World"
      Variable1
      
      Variable2 = "Goodbye World"
      Variable2
      
      Variable1 -> Variable2
      Variable1
      Variable2
```

*Removing a variable from Memory*
```{r}
rm(Variable1)
```

*Operators in R*
  *Relational operator:*
```{r}
#   a) Less than operator '<'
        3 < 4 
        4 < 3
#   b) Greater than operator '>'
        10 > 7
        7 > 10
#   c) Less than or equal to operator '<='
        8 <= 9
        9 <= 8
#   d) Greater than or equal to operator '>='
        9 >= 8
        8 >= 9
#   e) Equal to operator '=='
        1 == 1
        1 == 0
#   f) Not equal to operator '!='
        1 != 0
        1 != 1
```
 
  *Mathematical Operators:*
```{r}
#   a) Addition Operator '+'
          3 + 4 
#   b) Subtraction operator '-'
          10 - 7
#   c) Multiplication operator '*'
          2 * 9
#   d) Division operator '/'
          8 / 2
#   e) Eponent operator '^'
          2 ^ 4
#   f) Modulus than operator '%%'
          8 %% 7
```
  
  *Logical Operators:*
```{r}
#   a) And Operator '&'
          (8 > 7) & (6 == 6)
          (6 < 6) & (2 < 3)
#   b) Or operator '|'
          (6 < 6) | (2 < 3)
          (3 != 3) | (10 <= 8)
#   c) Not operator '!'
          !(8 > 7)
          !(6 < 6)
```
*Atomic Datatypes:*
```{r}
#   a) Character
        Var = 'Hello'
        typeof(Var)
#   b) Integer
        Var = 7L
        typeof(Var)
#   b) Numeric/Double
        Var = 3.14
        typeof(Var)
#   b) Complex
        Var = 7 + 2i
        typeof(Var)
#   b) Boolean/Logical
        Var = TRUE
        typeof(Var)
```       
 
        
*Conditionals in R*
 In order to make decisions selectively run code R uses conditional statements
    *If else block:*
```{r}
        if(Var){
          print("The condition is True")
        }else{
          print("The condition is False")
        }
```
    *else if block:*
        else if(!Var){
          print("The condiont is not True, its False")
        }
    *Switch Block:*
```{r}
        x = switch(3,
          "first",
          "second",
          "third",
          "fourth"
          )
        print(x)
``` 

*Loops in R*
Loops are used to repeatedly run the same block of code to perform some function
  *Repeat Loops:*'
```{r}
        counter = 1
        repeat{ 
          print(paste("Loop:",counter))
          counter = counter + 1
          
          if(counter == 4){
            break
          }
        }
```
  *While Loops:*
```{r}
        counter = 1
        while (counter < 4) {
          print(paste("Loop:",counter))
          counter = counter + 1
        }
```
  *For loops:*
```{r}
        for(counter in 1:4){
          print(paste("Loop:",counter))
        }
```

*Functions in R*
Functions are used to repeat the same set of code in different circumstances and different values, there are also several built-in functions which make programming much easier.eg: sqrt(), sum(), mean() etc.
```{r}
        func = function() {
          print("This is a function") 
        }
        func()
```

*Example Program:*
```{r}
        PrimeCheck = function(num){
          res = "It is a Prime Number"
          squareroot = ceiling(sqrt(num));
          if (num %% 2 != 0 && num != 1)
            for(i in seq(3,squareroot,2) )
            {
              if (num %% i == 0)
              {
                res = "It is not a Prime Number";
                break;
              }
            }
          else if(num == 2) {
            res = "It is a Prime Number";
          }
          else{
            res = "It is not a Prime Number";
          }
          return(res);
        }
        
        print(PrimeCheck(999))
```
        